master thesis

### 2018-03-14 Paper - Latex/Paper - Skizze - v3.tex
Paper - Methods - Ancestral state reconstruction methodes - Fitch Maximum Parsimony

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \section{ancestral state reconstruction methods}
      Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
      The first is to explain the current state with the least number of state changes between an 
        ancestor and his child, this is called parsimonious.
      The other class she presents involves modeling the character evolution as a stochastic process and 
        using the likelihoods to compute the possible ancestral character states. This is generally done 
        with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

@@

      Since the simpler parsimony methods are sufficient for our case, we have decided for them. \\
      \todotext{In the following we introduce ... there are these methods .. which are important for us ...}

      %---------------------------------------------------------------------------------------------------
      %---------------------------------------------------------------------------------------------------
      \subsection{Fitch maximum parsimony}
        Described from [COO98] + others ... - implemented for multifurcating trees

        Fitch algorithm for binary trees:
        \begin{figure}
          \caption{bla}
          \centering
            \includegraphics[width=0.5\textwidth]{Figures/Fitch1.png}
        \end{figure}

        Der Baum hat die folgende Struktur: Alle inneren Knoten sind leer. In den Blattknoten befindet 
        sich entweder das Tag FL oder P, oder deren Vereinigung, wenn es sich um einen unknown node handelt.

        Der Fitch Algorithmus ist aufgeteilt in drei Schritte, in welchen man jeweils durch den Baum traversiert.
        Schritt 1 beginnt von de Blättern aus, da sich dort zu Beginn die einzige Information befindet. 
        Für jeden Knoten gilt, wenn seine Kinder schon Information enthalten, dann bilde die Schnittmenge 
        der states und schreibe diese als Information in den aktuellen Knoten. Ist die Schnittmenge leer, 
        dann schreibe die Vereinigung aller möglichen states in den Knoten. Für alle Kinder, die noch keine 
        Information haben, führe diesen Schritt erst für diese aus.
        Schritt zwei geht von den Kindern der Wurzel bis zu den Vätern der Blätter. Jeder Knoten bekommt 
        einen zweiten Tag, der sich aus der Vereinigung des states des Vaterknoten und der Geschwisterknoten
        zusammensetzt. Ist diese leer, bekommt der Knoten wieder die Vereinigung aller states, also $\{FL,P\}$ als Tag.

        Hier gibt es einige Möglichkeiten, wie dieser Schritt genau aussieht.
        1. Version: Es wird nur der erste Tag vom Vaterknoten genutzt. Außerdem wird von den 
        Geschwisterknoten zuerst der Schnitt gebildet, und danach vom Ergebnis nochmal mit dem Vaterknoten zusammen.
        (Immer wenn der Schnitt leer ist, ist das Ergebnis die Vereinigung aller states, also $\{FL,P\}$. Auch im folgenden...)
        2. Version: Es wird nur der erste Tag vom Vaterknoten genutzt. Er wird zusammen mit den 
        Geschwisterstates genommen und direkt ein Schnitt aller Mengen gebildet.
        3. Version: Es werden alle vorherigen states vom Vater genutzt und von diesen ein Schnitt gebildet.
        Das selbe gilt für die Geschwisterstates. Und dann wird ein dritter Schnitt zwischend en Ergebnissen gebildet.
        4. Version: Es werden alle states genutzt und direkt in einem Schnitt zusammengenommen.

        Der Finale Schritt traversiert nochmal über den Baum und Bildet aus den zwei states pro Knoten einen finalen
        Tag, indem wieder der Schnitt der beiden states das Ergebnis ist.

@@

      %---------------------------------------------------------------------------------------------------
      \subsubsection{Sankoff}
        Maximum parsimony algorithm from Sankoff implemented in the R package castor \cite{Louca2017}.

@@

        \begin{table}[h]
          \begin{center}
              \begin{tabular}{ |l|r|r|r|r| }
                \hline
                Model / Taxa & Kingdom & Phylum & Class & Order \\
                \hline \hline
                multifurc $\sim$ taxa & 7774454 & \cellcolor{green!10}7435700 & \cellcolor{green!15}7337241 & \cellcolor{green!30}7076068 \\
                \hline
                multifurc $\sim$ taxa + depth & 7752303 & \cellcolor{green!10}7431609 & \cellcolor{green!15}7334754 & \cellcolor{green!30}7027578 \\
                multifurc $\sim$ taxa + max.height & 7730196 & \cellcolor{green!15}7375889 & \cellcolor{green!20}7275856 & \cellcolor{green!30}7005424 \\
                multifurc $\sim$ taxa + min.height & \cellcolor{green!10}7472500 & \cellcolor{green!20}7233486 & \cellcolor{green!25}7144686 & \cellcolor{green!40}6890703 \\
                multifurc $\sim$ taxa + mean.height & \cellcolor{green!15}7304402 & \cellcolor{green!25}7128318 & \cellcolor{green!30}7055313 & \cellcolor{green!40}6815271 \\
                \hline
                multifurc $\sim$ taxa * depth & 7714881 & \cellcolor{green!15}7335396 & \cellcolor{green!20}7250759 & \cellcolor{green!40}6843004 \\
                multifurc $\sim$ taxa * max.height & \cellcolor{green!5}7692980 & \cellcolor{green!15}7311241 & \cellcolor{green!25}7187504 & \cellcolor{green!45}6795823 \\
                multifurc $\sim$ taxa * min.height & \cellcolor{green!10}7442387 & \cellcolor{green!25}7177002 & \cellcolor{green!30}7094933 & \cellcolor{green!45}6795099 \\
                multifurc $\sim$ taxa * mean.height & \cellcolor{green!20}7247309 & \cellcolor{green!30}7020258 & \cellcolor{green!35}6965794 & \cellcolor{green!50}6665565 \\
                \hline
              \end{tabular}
          \end{center}
          \caption{Residuals...}
          \label{table:...} 
        \end{table}
        * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\

@@

        Because of the difference in the complexity of the models, we compared their BICs:
        \begin{table}[h]
          \begin{center}
            \begin{tabular}{ |l|r|r|r|r| }
              \hline
              Model / Taxa & Kingdom & Phylum & Class & Order \\
              \hline \hline
              multifurc $\sim$ taxa & 8273333 & \cellcolor{green!15}7937828 & \cellcolor{green!20}7842157 & \cellcolor{green!30}7644249\\
              \hline
              multifurc $\sim$ taxa + depth & 8273318 & \cellcolor{green!15}7934322 & \cellcolor{green!20}7839364 & \cellcolor{green!35}7539999 \\
              multifurc $\sim$ taxa + max.height & \cellcolor{green!15}7993515 & \cellcolor{green!25}7749121 & \cellcolor{green!30}7661817 & \cellcolor{green!40}7416211 \\
              multifurc $\sim$ taxa + min.height & 8251211 & \cellcolor{green!20}7875521  & \cellcolor{green!25}7778327 & \cellcolor{green!35}7516883 \\
              multifurc $\sim$ taxa + mean.height & \cellcolor{green!20}7825417 & \cellcolor{green!30}7644249 & \cellcolor{green!35}7572474 & \cellcolor{green!45}7340741 \\
              \hline
              multifurc $\sim$ taxa * depth & 8235932 & \cellcolor{green!20}7836755 & \cellcolor{green!25}7757688 & \cellcolor{green!45}7383808 \\
              multifurc $\sim$ taxa * max.height & \cellcolor{green!15}7963438 & \cellcolor{green!30}7693555 & \cellcolor{green!30}7614820 & \cellcolor{green!45}7335338 \\
              multifurc $\sim$ taxa * min.height & 8214030 & \cellcolor{green!20}7808940 & \cellcolor{green!30}7690618 & \cellcolor{green!45}7336627\\
              multifurc $\sim$ taxa * mean.height & \cellcolor{green!25}7768360 & \cellcolor{green!35}7536296 & \cellcolor{green!50}7484953 & \cellcolor{green!50}7206369 \\
              \hline
            \end{tabular} 
          \end{center}
          \caption{BIC...}
          \label{table:...} 
        \end{table}

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class \\
            \hline \hline
            multifurc $\sim$ taxa & 545799 & \cellcolor{green!10}500004 & \cellcolor{green!50}485121 \\
            \hline
            multifurc $\sim$ taxa + depth & 544862 & \cellcolor{green!30}493808 & \cellcolor{green!50}481869 \\
            \hline
            multifurc $\sim$ taxa * depth & 544179 & \cellcolor{green!50}489845 & \cellcolor{green!50}481494 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}

>>>>

    Multifurcation \\
    binary or bifurcating... \cite{Felsenstein2003}

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \section{Ancestral state reconstruction methods}
      For about 50 years, people have been working on ancestral state reconstruction. The first paper 
        mentioned above is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state 
        data \cite{Camin1965}. Different methods have been developed and the question is which method is 
        the most suitable for our problem. \\
      For this purpose, various studied methods and their advantages and disadvantages were compared. \\

      Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
      The first is to explain the current state with the least number of state changes between an 
        ancestor and his child, this is called parsimonious. \\
      The other class she presents involves modeling the character evolution as a stochastic process and 
        using the likelihoods to compute the possible ancestral character states. This is generally done 
        with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

@@

      Following the principle of the simpler model first, the decision has \todo{fallen?} on parsimonious 
        methods since these are sufficient for the case present here. \\
      Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
        previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
        \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
      \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
        Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
      \todo{Wagner-parsimony \cite{Swofford1987}} \\

      Thus, the methods used in this work are those of Fitch and Sankoff. For Fitch, the algorithm has 
        been extended from binary to multifurcated trees. For the Sankoff algorithm, Louca and Doebeli 
        have presented an implementation for non-binary trees published in an R package named 
        \textit{castor} \cite{Louca2017}.

      %---------------------------------------------------------------------------------------------------
      %---------------------------------------------------------------------------------------------------
      \subsection{Fitch maximum parsimony}
        Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
          reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
        Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
          and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
          simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
          these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf multifurcated angepasst...} \\
        To understand the differences to the multifurcated case, the algorithm for the binary case is 
          explained briefly. \\

        Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
          a set of states. Internal nodes have three sets, which are empty at the beginning, excluding the
          root node, it has only one. Leaf nodes have their state as a set (eg \{FL\}, unknown leaf 
          nodes) the union of all possible states (\{FL, P\}). \\
        The algorithm traverses three times trough the tree and fills these sets. \\
        In each step, two sets are considered and their intersect formed. There are two cases:
        \begin{wrapfigure}{r}{0.4\textwidth}
          \centering
          \includegraphics[width=0.39\textwidth]{Figures/Fitch1.png}
          \caption{Fitch algorithm for binary trees}
          \label{fig: binary Fitch}
        \end{wrapfigure}
        \begin{enumerate}
          \item The intersection is not empty and corresponds to the new set.
          \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
        \end{enumerate}
        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal. Each internal node is formed from its child nodes. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter \todotext{does it? Welche Mengen/Menge wird genommen 
          von jedem Knoten?}): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\

        Figure \ref{fig: binary Fitch} shows the firch algorithm for binary trees with 5 leaf nodes, \todo{one unknown}.

        The finale step is not part of the original Fitch algorithm, as it only aims to minimize the 
          transitions without predicting actual states of internal nodes. This step is described by \todo{...}. \\

        % Der Fitch Algorithmus ist aufgeteilt in drei Schritte, in welchen man jeweils durch den Baum traversiert.
        % Schritt 1 beginnt von de Blättern aus, da sich dort zu Beginn die einzige Information befindet. 
        % Für jeden Knoten gilt, wenn seine Kinder schon Information enthalten, dann bilde die Schnittmenge 
        % der states und schreibe diese als Information in den aktuellen Knoten. Ist die Schnittmenge leer, 
        % dann schreibe die Vereinigung aller möglichen states in den Knoten. Für alle Kinder, die noch keine 
        % Information haben, führe diesen Schritt erst für diese aus.
        % Schritt zwei geht von den Kindern der Wurzel bis zu den Vätern der Blätter. Jeder Knoten bekommt 
        % einen zweiten Tag, der sich aus der Vereinigung des states des Vaterknoten und der Geschwisterknoten
        % zusammensetzt. Ist diese leer, bekommt der Knoten wieder die Vereinigung aller states, also $\{FL,P\}$ als Tag.

        % Hier gibt es einige Möglichkeiten, wie dieser Schritt genau aussieht.
        % 1. Version: Es wird nur der erste Tag vom Vaterknoten genutzt. Außerdem wird von den 
        % Geschwisterknoten zuerst der Schnitt gebildet, und danach vom Ergebnis nochmal mit dem Vaterknoten zusammen.
        % (Immer wenn der Schnitt leer ist, ist das Ergebnis die Vereinigung aller states, also $\{FL,P\}$. Auch im folgenden...)
        % 2. Version: Es wird nur der erste Tag vom Vaterknoten genutzt. Er wird zusammen mit den 
        % Geschwisterstates genommen und direkt ein Schnitt aller Mengen gebildet.
        % 3. Version: Es werden alle vorherigen states vom Vater genutzt und von diesen ein Schnitt gebildet.
        % Das selbe gilt für die Geschwisterstates. Und dann wird ein dritter Schnitt zwischend en Ergebnissen gebildet.
        % 4. Version: Es werden alle states genutzt und direkt in einem Schnitt zusammengenommen.

        % Der Finale Schritt traversiert nochmal über den Baum und Bildet aus den zwei states pro Knoten einen finalen
        % Tag, indem wieder der Schnitt der beiden states das Ergebnis ist.

@@

      %---------------------------------------------------------------------------------------------------
      \subsubsection{Sankoff}
        Maximum parsimony algorithm from Sankoff implemented in the R package castor \cite{Louca2017}. \\
        \todo{transition probabilities: all equal}

@@

        \begin{table}[h]
          \begin{center}
              \begin{tabular}{ |l|r|r|r|r|r| }
                \hline
                Model / Taxa & Kingdom & Phylum & Class & Order & Family \\
                \hline \hline
                multifurc $\sim$ taxa & 7774454 & \cellcolor{green!10}7435700 & \cellcolor{green!15}7337241 & \cellcolor{green!30}7076068 \\
                \hline
                multifurc $\sim$ taxa + depth & 7752303 & \cellcolor{green!10}7431609 & \cellcolor{green!15}7334754 & \cellcolor{green!30}7027578 \\
                multifurc $\sim$ taxa + max.height & 7730196 & \cellcolor{green!15}7375889 & \cellcolor{green!20}7275856 & \cellcolor{green!30}7005424 \\
                multifurc $\sim$ taxa + min.height & \cellcolor{green!10}7472500 & \cellcolor{green!20}7233486 & \cellcolor{green!25}7144686 & \cellcolor{green!40}6890703 \\
                multifurc $\sim$ taxa + mean.height & \cellcolor{green!15}7304402 & \cellcolor{green!25}7128318 & \cellcolor{green!30}7055313 & \cellcolor{green!40}6815271 \\
                \hline
                multifurc $\sim$ taxa * depth & 7714881 & \cellcolor{green!15}7335396 & \cellcolor{green!20}7250759 & \cellcolor{green!40}6843004 \\
                multifurc $\sim$ taxa * max.height & \cellcolor{green!5}7692980 & \cellcolor{green!15}7311241 & \cellcolor{green!25}7187504 & \cellcolor{green!45}6795823 \\
                multifurc $\sim$ taxa * min.height & \cellcolor{green!10}7442387 & \cellcolor{green!25}7177002 & \cellcolor{green!30}7094933 & \cellcolor{green!45}6795099 \\
                multifurc $\sim$ taxa * mean.height & \cellcolor{green!20}7247309 & \cellcolor{green!30}7020258 & \cellcolor{green!35}6965794 & \cellcolor{green!50}6665565 \\
                \hline
              \end{tabular}
          \end{center}
          \caption{Residuals...}
          \label{table:...} 
        \end{table}
        * Residuals: Fehler - wieviele Werte sind nicht gut modelliert. (umso kleiner umso besser - grün) \\

@@

        Because of the difference in the complexity of the models, we compared their BICs:
        \begin{table}[h]
          \begin{center}
            \begin{tabular}{ |l|r|r|r|r|r| }
              \hline
              Model / Taxa & Kingdom & Phylum & Class & Order & Family \\
              \hline \hline
              multifurc $\sim$ taxa & 8273333 & \cellcolor{green!15}7937828 & \cellcolor{green!20}7842157 & \cellcolor{green!30}7644249 \\
              multifurc $\sim$ taxa & 8257680 & \cellcolor{green!15}7922207 & \cellcolor{green!20}7826490 & \cellcolor{green!35}7574154 \\
              \hline
              multifurc $\sim$ taxa + depth & 8273318 & \cellcolor{green!15}7934322 & \cellcolor{green!20}7839364 & \cellcolor{green!35}7539999 \\
              multifurc $\sim$ taxa + max.height & \cellcolor{green!15}7993515 & \cellcolor{green!25}7749121 & \cellcolor{green!30}7661817 & \cellcolor{green!40}7416211 \\
              multifurc $\sim$ taxa + min.height & 8251211 & \cellcolor{green!20}7875521  & \cellcolor{green!25}7778327 & \cellcolor{green!35}7516883 \\
              multifurc $\sim$ taxa + mean.height & \cellcolor{green!20}7825417 & \cellcolor{green!30}7644249 & \cellcolor{green!35}7572474 & \cellcolor{green!45}7340741 \\
              \hline
              multifurc $\sim$ taxa * depth & 8235932 & \cellcolor{green!20}7836755 & \cellcolor{green!25}7757688 & \cellcolor{green!45}7383808 \\
              multifurc $\sim$ taxa * max.height & \cellcolor{green!15}7963438 & \cellcolor{green!30}7693555 & \cellcolor{green!30}7614820 & \cellcolor{green!45}7335338 \\
              multifurc $\sim$ taxa * min.height & 8214030 & \cellcolor{green!20}7808940 & \cellcolor{green!30}7690618 & \cellcolor{green!45}7336627\\
              multifurc $\sim$ taxa * mean.height & \cellcolor{green!25}7768360 & \cellcolor{green!35}7536296 & \cellcolor{green!50}7484953 & \cellcolor{green!50}7206369 \\
              \hline
            \end{tabular} 
          \end{center}
          \caption{BIC...}
          \label{table:...} 
        \end{table}

@@

      \begin{table}[h!]
        \begin{center}
          \begin{tabular}{ |l|r|r|r|r| }
            \hline
            Model / Taxa & Kingdom & Phylum & Class & Order \\
            \hline \hline
            multifurc $\sim$ taxa & 545799 & \cellcolor{green!10}500004 & \cellcolor{green!50}485121 \\
            \hline
            multifurc $\sim$ taxa + depth & 544862 & \cellcolor{green!30}493808 & \cellcolor{green!50}481869 & 478851 \\
            \hline
            multifurc $\sim$ taxa * depth & 544179 & \cellcolor{green!50}489845 & \cellcolor{green!50}481494 \\
            \hline
          \end{tabular} 
        \end{center}
        \caption{BICs of unknown information}
        \label{table:BIC unknown information} 
      \end{table}
